Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-19 aistats.tex
Formalizer  theorem for multivariate MLE

{\bf Duality.}
The log-partition function $\logpart$ verifies the two following identities:
\begin{align}
    \nabla\logpart(\nat) &=  \expect[p_\nat]{T(X)} =: \meanp, \\
    \nabla^2 \logpart(\nat) &= \Cov_{p_\nat}[T(X)] > 0,
\end{align}
where $\meanp$ is called the mean (or dual) parameter, which lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
Given that $\logpart$ is strictly convex, its Hessian is positive definite, and its gradient $\nabla \logpart$ is a \textit{bijection} between natural parameters $\nat$ and mean parameters $\m$.
We will write $\m$ or $\nat$ interchangeably depending on the context, being aware that both are linked and represent the same distribution.

@@

	\begin{theorem}[MLE Upper Bound]
	The MLE of $\cN(0,\mu_*)$ is $\mu_n = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$ and otherwise
	\begin{align}
		 \expect{\bregmanconj( \mu_*; \mu_n) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
	\end{align}
	\end{theorem}

@@

The entropy of $X$ is a negative log-determinant $\conj(\mu) = -\log\det(\mu)$, whose gradient is the negative matrix inverse $\nabla\conj (\mu) = - \mu^{-1}$.
The associated Bregman divergence is
\alignn{
	\bregmanconj(\mu_*; \hat \mu_n) 
	= \half \paren{\Tr(\mu_* \hat \mu_n^{-1} ) - d  - \log\det(\mu_* \hat \mu_n^{-1}) } \; .
}
Thanks to the linearity of the trace, the expectation becomes
\alignn{
	\expect{\bregmanconj(\mu_*; \hat \mu_n) }
	= \half \paren{\Tr(\expect{\mu_* \hat \mu_n^{-1}} ) - d   
	+\expect{\log\det( \hat \mu_n \mu_*^{ -1}))} } \; .
}

When the estimator is the MLE $\hat \mu_n = \inv{n} \sum_i X_i X_i^\top$, 
then the mean parameter "ratio" is
\aligns{
	\mV = n \mu_*^{-\half}\hat \mu_n \mu_*^{-\half} = \sum_i (\mu_*^{-\half} X_i) (\mu_*^{-\half}X_i)^\top
}
but $\mu_*^{-\half}X_i \sim \cN(0,\mI)$, so that $\mV$
is sampled from a Wishart  $\cW(\mI, n)$, where $\mI$ stands for the identity matrix of order $d$. 
We know formulas for the expectation of an inverse Wishart and the log-determinant of a Wishart (thanks to $\log\det$ being a sufficient statistics and $\nabla\logpart$)
\aligns{
	\E[\mV] &= n \mI \\
	\forall n > d+1,\E[\mV^{-1}] &= \frac{\mI}{n - d - 1} 
	\implies \Tr(\E[\mV^{-1}]) = \frac{d}{n - d - 1}  \\
	\E[\log\det \mV] &= \psi_d(\half[n]) + d \log 2 \; ,
}
where $\psi_d(\half[n]) = \sum_{i=0}^{d-1} \psi ( \half[n - i])$ is the multivariate digamma function.
Putting it all together, we get the following closed form for the expectation of the KL
 \begin{align}
 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
 	= \half \left ( \frac{d(d+1)}{n-d-1} + \psi_d(\half[n]) - d \log(\half[n])  \right )
 \end{align}

The complicated term is the $\psi_d$ function.
We can use the upper bound $\psi(x) - \log(x) \leq - \inv{2x}$ (like in the univariate case) to get
\begin{align}
	\psi_d(\half[n]) - d \log(\half[n]) 
	&= \sum_{i=0}^{d-1} \paren{\psi ( \half[n - i]) - \log(\half[n]) } \\
	&\leq \sum_{i=0}^{d-1} \paren{\log(1 - \frac{i}{n}) - \frac{1}{n-i}} \; .
	\label{eq:psid_bound}
\end{align}
We can bound the harmonic sum thanks to the typical bound
\aligns{
	\inv{2n +1}
	\leq H_n - \log(n) - \gamma \leq 
	\inv{2n - 1}
}
where $\gamma$ is the Euler constant. Then
\begin{align}
	- \sum_{i=0}^{d-1} \frac{1}{n-i} 
	&= H_{n-d} - H_n \\
	&\leq \log(n-d) + \gamma + \inv{2(n-d) - 1} - \log(n) - \gamma - \inv{2n+1}\\
	&= \log(1-\frac{d}{n})  + \frac{2(d + 1)}{(2n+1)(2(n - d) - 1}
\end{align}
Plugging this back into~\eqref{eq:psid_bound} yields
\begin{align}
	\psi_d(\half[n]) - d \log(\half[n]) 
	\leq \sum_{i=0}^{\textcolor{red}{d}} \log(1 - \frac{i}{n})
	+ \frac{2(d + 1)}{(2n+1)(2(n - d) - 1} 
\end{align}
where we emphasized that the sum runs up to $d$.
Now to bound this sum of logarithms, we can use a concave tangent bound on $\log$ to get
\begin{align}
	\sum_{i=0}^{d} \log(1 - \frac{i}{n})
	\leq \sum_{i=0}^{d} - \frac{i}{n}
	 = -\frac{d(d+1)}{2n}
\end{align}
which yields the final rate
\begin{align}
 	\boxed{
 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
 	\leq \frac{p}{2n} + \frac{(d+1) \paren{p + \inv{4}}}{n (n - d -1)} 
 	\ , \forall n >d+1 \;.
 	}
\end{align}
where $p = d(d+1)/2$.
This rate is asymptotically tight.

>>>>

{\bf Duality.}
The log-partition function $\logpart$ verifies the two following identities:
\begin{align}
    \nabla\logpart(\nat) &=  \expect[p_\nat]{T(X)} =: \meanp, \label{eq:mirror-map} \\
    \nabla^2 \logpart(\nat) &= \Cov_{p_\nat}[T(X)] > 0,
\end{align}
where $\meanp$ is called the mean (or dual) parameter, which lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
Given that $\logpart$ is strictly convex, its Hessian is positive definite, and its gradient $\nabla \logpart$ is a \textit{bijection} between natural parameters $\nat$ and mean parameters $\m$.
We will write $\m$ or $\nat$ interchangeably depending on the context, being aware that both are linked and represent the same distribution.

@@

	\begin{theorem}[MLE Upper Bound]
	The MLE of $\cN(0,\mu_*)$ is $\mu_n = \inv{n} \sum_i X_i^2 $.
	The expected KL between $\mu_*$ and $\hat\mu_n$	is infinite when $n\leq 2$ and otherwise upper bounded by
	\begin{align}
		 \expect{\bregmanconj( \mu_*; \mu_n) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
	\end{align}
	\end{theorem}

@@

\begin{theorem}[Multivariate MLE Upper Bound]
	The MLE of  the covariance matrix of $X_i\sim\cN(0,\mu_*)$ is $\hat \mu_n = \inv{n} \sum_i X_i X_i^\top \in\real^{d\times d}$ with $p = \frac{d(d+1)}{2} $ degrees of freedom. 
	The expected KL between $\mu_*$ and $\hat\mu_n$ is infinite when $n\leq d+1$ and otherwise upper bounded by
	\alignn{
	 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
 	\leq \frac{p}{2n} + \frac{(d+1) \paren{p + \inv{4}}}{n (n - d -1)} 
 	\ , \forall n >d+1 \;.
}
This bound is asymptotically tight.
\end{theorem}
\begin{proof}
The entropy of $X$ is a negative log-determinant $\conj(\mu) = -\log\det(\mu)$, whose gradient is the negative matrix inverse $\nabla\conj (\mu) = - \mu^{-1}$.
The associated Bregman divergence is
\alignn{
	\bregmanconj(\mu_*; \hat \mu_n) 
	= \half \paren{\Tr(\mu_* \hat \mu_n^{-1} ) - d  - \log\det(\mu_* \hat \mu_n^{-1}) } \; .
}
Thanks to the linearity of the trace, the expectation becomes
\alignn{
	\expect{\bregmanconj(\mu_*; \hat \mu_n) }
	= \half \paren{\Tr(\expect{\mu_* \hat \mu_n^{-1}} ) - d   
	+\expect{\log\det( \hat \mu_n \mu_*^{ -1}))} } \; .
}

When the estimator is the MLE $\hat \mu_n = \inv{n} \sum_i X_i X_i^\top$, 
then the mean parameter "ratio" is
\aligns{
	\mV = n \mu_*^{-\half}\hat \mu_n \mu_*^{-\half} = \sum_i (\mu_*^{-\half} X_i) (\mu_*^{-\half}X_i)^\top
}
but $\mu_*^{-\half}X_i \sim \cN(0,\mI)$, so that $\mV$
is sampled from a Wishart  $\cW(\mI, n)$, where $\mI$ stands for the identity matrix of order $d$.
Recall that $\E[\mV] = n \mI$.
Thanks to $\log\det$ being a sufficient statistic of the Wishart, and the natural to mean parameter formula~\eqref{eq:mirror-map}, we have a closed form for the expected log-determinant of a Wishart 
\aligns{
	\E[\log\det \mV] &= \psi_d(\half[n]) + d \log 2 \; ,
}
where $\psi_d(\half[n]) = \sum_{i=0}^{d-1} \psi ( \half[n - i])$ is the multivariate digamma function.
The expectation of an inverse Wishart  is straightforward to compute from the density and the log-partition function
\begin{equation}
	\E[\mV^{-1}] = \begin{cases}
		+\infty \text{ if } n\leq d+1 \\
		 \frac{\mI}{n - d - 1} \text{ otherwise,} 
	\end{cases}
\end{equation}
which proves the infinite part of the statement.
We now consider only $n>d+1$.
We deduce that $\Tr(\E[\mV^{-1}]) = \frac{d}{n - d - 1}$.
Putting it all together, we get the following closed form for the expectation of the KL
 \begin{align}
 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
 	= \half \left ( \frac{d(d+1)}{n-d-1} + \psi_d(\half[n]) - d \log(\half[n])  \right )
 \end{align}

The complicated term is the $\psi_d$ function.
We can use the upper bound $\psi(x) - \log(x) \leq - \inv{2x}$ (like in the univariate case) to get
\begin{align}
	\psi_d(\half[n]) - d \log(\half[n]) 
	&= \sum_{i=0}^{d-1} \paren{\psi ( \half[n - i]) - \log(\half[n]) } \\
	&\leq \sum_{i=0}^{d-1} \paren{\log(1 - \frac{i}{n}) - \frac{1}{n-i}} \; .
	\label{eq:psid_bound}
\end{align}
We can bound the harmonic sum thanks to the typical bound
\aligns{
	\inv{2n +1}
	\leq H_n - \log(n) - \gamma \leq 
	\inv{2n - 1}
}
where $\gamma$ is the Euler constant. Then
\begin{align}
	- \sum_{i=0}^{d-1} \frac{1}{n-i} 
	&= H_{n-d} - H_n \\
	&\leq \log(n-d) + \gamma + \inv{2(n-d) - 1} - \log(n) - \gamma - \inv{2n+1}\\
	&= \log(1-\frac{d}{n})  + \frac{2(d + 1)}{(2n+1)(2(n - d) - 1}
\end{align}
Plugging this back into~\eqref{eq:psid_bound} yields
\begin{align}
	\psi_d(\half[n]) - d \log(\half[n]) 
	\leq \sum_{i=0}^{\textcolor{red}{d}} \log(1 - \frac{i}{n})
	+ \frac{2(d + 1)}{(2n+1)(2(n - d) - 1} 
\end{align}
where we emphasized that the sum runs up to $d$.
Now to bound this sum of logarithms, we can use a concave tangent bound on $\log$ to get
\begin{align}
	\sum_{i=0}^{d} \log(1 - \frac{i}{n})
	\leq \sum_{i=0}^{d} - \frac{i}{n}
	 = -\frac{d(d+1)}{2n}
\end{align}
which yields the final rate
\begin{align}
 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
 	\leq \frac{p}{2n} + \frac{(d+1) \paren{p + \inv{4}}}{n (n - d -1)} 
 	\ , \forall n >d+1 \;.
\end{align}
where $p = d(d+1)/2$.
\end{proof}
